An assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2019-12-11 manuscript/manuscript.tex
Backup before aggressively refactoring code in light of Kyle's comments

A large number of studies have shown relationships between biodiversity and ecosystem functionality \citep{Liang2016, Cardinale2009}. The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect aggregate measures of function such as resistance of productivity to disturbance, and even to static properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others functional contribution.

The relationship between biodiversity and ecosystem functionality has informed conservation policy which aims to maintain and strengthen biodiversity under the assumption that this strengthens ecosystem resilience to stressors such as anthropogenic climate change and ecosystem service use by humans \citep{Thompson2011}. It is probable however, that the strength and relative importance of the BEFR varies with ecosystem studied and environmental conditions, with many ecosystems remaining understudied in the BEFR literature.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which are unique in terms of species assemblage, being an oceanic island with many endemic species. In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity \citep{Wright2017, Poorter2015, Sande2017a}. In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as great a role in the relationship between biodiversity and ecosystem function. Instead, stress tolerance and the functional contribution of abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{Ratcliffe2017}.

Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{plot_loc}). The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}. Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}. Southern African woodlands have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity. Southern African woodlands are also relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}. This has raised interest in how biodiversity influences ecosystem function in these ecosystems. Resource extraction by humans is directly influencing biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Aleman2016, Ryan2016}. Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}. While rapid biodiversity change is being observed in southern African woodlands, research into the relationship between biodiversity and ecosystem functionality remains scarce.

A small number of academic studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}, but the scientific impact of all these studies has been hampered by a restricted climatic and biogeographical range of study sites. Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass, it is important to sample across these gradients to be able to infer a truly regional scale relationship between diversity and biomass. Studies conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Small scale studies may therefore be over-estimating the effect of diversity.

@@

While southern African woodlands are species rich in the herbaceous understorey \citep{Murphy2016}, the tree layer is relatively species poor. \citet{Solbrig1996} however, writes that they remain structurally diverse in the tree layer, with trees occupying distinct layers of the canopy at different growth stages and between species. This structural diversity may be one mechanism through which diversity influences woody biomass. \citep{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production. Tree species in southern African woodlands may differ in their occupation of different canopy layers allowing a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area as a form of niche complementarity. This theory however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not occur in savannas, which are structured by disturbance rather than competition. Instead, disturbance history may override the effects of species diversity on structural diversity nullifying the effects of species diversity on structural diversity.

\citep{Zhang2012} conducted a meta-analysis of 54 studies of the BEFR in forest ecosystems and found that greater tree species abundance evenness led to an increased effect size of species richness on productivity. They suggest this effect is caused by an increased heterogeneity of functional traits over space at small scales, leading to a more evenly filled environmental niche volume, which allows a greater number of species to maximise their productivity. We expect that this effect will occur in southern African woodlands as well. A decrease in species abundance evenness may be found as a result of increased disturbance, which prevents dominance by one species, or due to more extreme environmental conditions which preclude the growth of many species, instead favouring those few species who can effective tolerate sub-optimal conditions. It is important to therefore to consider not only species richness as a measure of diversity, but also the evenness of the distribution of trees among species when assessing the impact of diversity on ecosystem function.ecosystem function.

@@

\begin{enumerate}
	\item{Plots with a higher tree species diversity will maintain higher above-ground woody biomass stocks.}
	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species diversity on above-ground woody biomass.} 
	\item{There will be a positive effect of precipitation and soil fertility on AGB. Part of this positive effect will exist as a mediating effect through tree species diversity.}
	\item{Tree structural diversity of the woodland will interact with tree species diversity to provide an indirect path of influence between species diversity and woody biomass stocks.}
	\item{The observed effect of diversity on AGB will increase in strength as stem density increases owing to an increased importance of niche complementarity as competition increases.}
\end{enumerate}

@@

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1987}). The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The study sites straddle the climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}). The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

\begin{figure}[H]
	\centering
	\subfloat[]{{\includegraphics[width=0.3\textwidth]{plot_loc}}\label{plot_loc}}%
    \qquad
\subfloat[]{{\includegraphics[width=0.4\textwidth]{temp_precip_hull}}\label{temp_precip_hull}}%
\caption{The locations of the \nplots{} plots used in this study, by geographic location (a) with respect to the distribution of Miombo woodland vegetation according to \citet{White1987}, and in climate space (b), showing the plot locations as points compared to the climate space of the whole region as estimated using the WorldClim dataset over the Miombo woodland vegetation extent \citep{Fick2017}. Note that the density colour scale is log-transformed.}
\end{figure}

@@

Many plots provided by the Zambian Forestry Commission were arranged in clusters of up to four 20x50 m plots, 20 metres apart. Plots within a cluster were aggregated before the plot dataset filtering described above and treated as a single plot in analyses.

After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition. We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbor criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers. We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, removing \noutliers{} plots.

@@

We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like but excluding lianas, which fill a different ecological niche \citep{Selaya2008}. Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. Most plots in the dataset did not include data on stems <5 cm DBH, with these small stems comprising a very small proportion of the total AGB in a plot.

@@

Climatic data was collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}. Values of Mean Annual Temperature (MAT) and Mean Annual Precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature and precipitation seasonality were both calculated as the coefficient of variation of daily MAT and MAP, respectively, across the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), percentage soil organic carbon by volume (Org. C \%), and percentage soil sand content by volume (Sand \%). These data are a modelled product compiled from various remote sensed and directly measured data sources.

@@

\subsection{Data analysis}
Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure uses Hill numbers to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness was preferred over raw Hill numbers as they are more interpretable, representing actual species numbers. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.

@@

Plots were assigned to vegetation type groups based on tree species composition. Groups were identified in \citet{Fayolle2018} using an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data. Groups were identified using unconstrained correspondence analysis and ordination. Plot data used in this study occurred in five vegetation type groups. See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters .

@@

Structural Equation Models (SEM) investigated the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is impossible in multiple regression. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset. For example, moisture availability is expected to affect AGB \citep{Saito2014, Campbell1996}, but moisture availability itself is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration. Independently of total precipitation, precipitation seasonality determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth. Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and woody species diversity, which could otherwise increase the chances of Type I error and wrongly attributing inference due to covariance of explanatory variables, if testing for significant effects of biodiversity on AGB using conventional regression analyses \citep{Nachtigall2003}.

@@

First, we assessed the interacting factors of structural diversity and species diversity in determining AGB. We constructed a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately. We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a path with other paths of inference held constant. Model fit was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.

@@

The direction of these correlations was used as a test of our assumptions of the direction of influence of of latent variables later used in the SEMs. As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a negative correlation between the seasonality of precipitation and AGB (\ccmcb{}. MAT and the seasonality of temperature negatively correlated with AGB (MAT: \cctb{}; MAT seasonality: \cctcb{}). As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil organic Carbon and AGB (\ccob{}).

MAP had positive correlations with tree species richness (\ccms{}), abundance evenness (\ccme{}), tree height diversity (\ccmh{}) and a positive but not significant correlation with tree stem density. MAT had non significant correlations with tree species and structural diversity variables. Tree species diversity variables had clear positive correlations with stem density (Species richness: \ccsi{}; Shannon equitability: \ccei{}).

@@

Cluster 2, which contained vegetation dominated by \textit{Julbernadia paniculata} showed no significant effect of tree species diversity on AGB (\strucbsb{}), in contrast to the other vegetation clusters, and had a significantly smaller positive effect of structural diversity on AGB (\strucbhb{}) (\autoref{struc_model_slopes_all}).

@@

We repeatedly sub-sampled the plot dataset to build \subn{} datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density. Each dataset consisted of approximately \subp{} plots with of plot identity between subsampled datasets. The same SEM specification was used as above (\autoref{struc_mod}) except that stem density was removed.  \autoref{sem_struc_stems_ha} shows a saturating effect of tree species diversity on AGB as stem density increases. There appears to be a minimum stem density threshold at \textapprox{}295 stems ha\textsuperscript{-1} below which there appears to be no effect of tree diversity on biomass. The effect of structural diversity on biomass appears to decrease slightly with increasing stem density, though the relationship is less clear.

@@

Similar to the model which only considered tree species and structural diversity, the direct effect of species diversity on structural diversity was very weakly positive, while structural diversity itself had a positive effect on AGB (\rghb{}).

@@

Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass (\autoref{int_plots}). The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant (\moderp{}), while the interaction term of tree species diversity and soil fertility was not (\moders{}). As moisture availability increased, the effect of tree species diversity on AGB became stronger. See \autoref{mois_div_int_mod} and \autoref{soil_div_int_mod} for full regression tables.

@@

\bibliography{/Users/johngodlee/google_drive/bib/lib.bib}

>>>>

A large number of studies have shown relationships between biodiversity and ecosystem functionality (e.g. \citealt{Liang2016, Cardinale2009}). The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and even to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others functional contribution.

%The relationship between biodiversity and ecosystem functionality has informed conservation policy which aims to maintain and strengthen biodiversity under the assumption that this strengthens ecosystem resilience to stressors such as anthropogenic climate change and ecosystem service use by humans \citep{Thompson2011}. It is probable however, that the strength and relative importance of the BEFR varies with ecosystem studied and environmental conditions, with many ecosystems remaining understudied in the BEFR literature.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages. In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity \citep{Wright2017, Poorter2015, Sande2017a}. In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as great a role in the relationship between biodiversity and ecosystem function. Disturbance may weaken the role of competition in determining local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979,Grace1990}. Instead, stress tolerance and the functional contribution of mor abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{Ratcliffe2017}.

Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{plot_loc}). The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}. Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}. Southern African woodlands have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity.

Southern African woodlands are also relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}. This has raised interest in how biodiversity influences ecosystem function in these ecosystems. Resource extraction by humans is directly influencing biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Aleman2016, Ryan2016}. Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}. While rapid biodiversity change is being observed in southern African woodlands \citep{}, research into the relationship between biodiversity and ecosystem functionality remains scarce.

A small number of studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}, but the scientific impact of these studies has been hampered by a restricted climatic and biogeographical range of study sites. Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass, it is important to sample across these gradients to be able to infer a truly regional scale relationship between diversity and biomass. Studies conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Small scale studies may therefore be over-estimating the effect of diversity.

@@

While southern African woodlands are species rich in the herbaceous understorey \citep{Murphy2016}, the tree layer is relatively species poor. \citet{Solbrig1996} however, writes that they remain structurally diverse in the tree layer, with trees occupying distinct layers of the canopy at different growth stages and among species. This structural diversity may be one mechanism through which diversity influences woody biomass. \citep{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production. Tree species in southern African woodlands may differ in their occupation of different canopy layers allowing a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area as a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition. Instead, disturbance history may override the effects of species diversity on structural diversity nullifying the effects of species diversity on structural diversity.

\citep{Zhang2012} conducted a meta-analysis of 54 studies of the BEFR in forest ecosystems and found that greater tree species abundance evenness led to an increased effect size of species richness on productivity. They suggest this effect is caused by an increased heterogeneity of functional traits over space at small scales, leading to a more evenly filled environmental niche volume, which allows a greater number of species to maximise their productivity. We expect that this effect will occur in southern African woodlands as well. A decrease in species abundance evenness may be found as a result of increased disturbance, which prevents dominance by one species, or due to more extreme environmental conditions which preclude the growth of many species, instead favouring those few species which can effectively tolerate sub-optimal conditions. It is important therefore to consider not only species richness as a measure of diversity, but also the evenness of the distribution of trees among species when assessing the impact of diversity on ecosystem function.

@@

\begin{enumerate}
	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species diversity on above-ground woody biomass.} 
	\item{Precipitation and soil fertility will indirectly positively effect AGB via an increase in tree species diversity.}
	\item{Tree structural diversity will indirectly influence tree species diversity to provide an indirect path of influence between structural diversity and AGB.}
	\item{The effect size of species diversity on AGB will increase with stem density, due to an increased importance of niche complementarity as competition increases.}
\end{enumerate}

@@

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1987}). The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The study sites span the core climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}). The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.3\textwidth]{plot_loc}}\label{plot_loc}}%
	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of Miombo woodland vegetation according to \citet{White1987}. Points are coloured according to their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
\end{figure}

@@

Many plots provided by the Zambian Forestry Commission were arranged in clusters of up to four 20x50 m plots, 20 metres apart. Plots within each cluster were aggregated before the plot dataset filtering described above and treated as a single plot in analyses.

After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition. We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbor criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers. We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, thusremoving \noutliers{} plots.

@@

We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like, but excluding lianas, which fill a different ecological niche \citep{Selaya2008}. Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. Most plots in the dataset did not include data on stems <5 cm DBH. For those plots with stem measurements <5 cm DBH, those small stems only accounted for a median average of 0.42\% of the plot level AGB.

@@

Climatic data were collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}. Values of Mean Annual Temperature (MAT) and Mean Annual Precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature and precipitation seasonality were both calculated as the coefficient of variation of daily MAT and MAP, respectively, across the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), percentage soil organic carbon by volume (Org. C \%), and percentage soil sand content by volume (Sand \%). These data are a modelled product derivedfrom various remotel sensed and directly measured data sources.

@@

\subsection{Data analysis}
Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.

@@

Plots were assigned to vegetation type groups based on tree species composition. Groups were identified in \citet{Fayolle2018} via an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data. Groups were identified using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes. Plot data used in this study occurred in five vegetation type groups. See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters .

@@

Structural Equation Models (SEM) investigated the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is challenging in standard multiple regression. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset. For example, moisture availability is expected to affect AGB \citep{Saito2014, Campbell1996}, but moisture availability itself is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation, temperature which affects the rate of evapotranspiration and soil water retention. Independently of total precipitation, precipitation seasonality determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth. Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and woody species diversity, which could otherwise increase the chances of Type I error and wrongly attributing inference due to covariance of explanatory variables, if testing for significant effects of biodiversity on AGB using conventional regression analyses \citep{Nachtigall2003}.

@@

First, we assessed the interacting factors of structural diversity and species diversity in determining AGB. We constructed a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately. We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a path with other paths of inference held constant. Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.9, TLI >0.9, RMSEA <0.1, alongside our judgement of the model estimates.

@@

The direction of these correlations was used as a test of our assumptions of the direction of influence of latent variables later used in the SEMs. As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a negative correlation between the seasonality of precipitation and AGB (\ccmcb{}. MAT and the seasonality of temperature negatively correlated weakly with AGB (MAT: \cctb{}; MAT seasonality: \cctcb{}). As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil organic carbon and AGB (\ccob{}).

MAP had positive correlations with tree species richness (\ccms{}), abundance evenness (\ccme{}), tree height diversity (\ccmh{}) and a positive but not significant correlation with tree stem density. MAT had weak correlations with tree species and structural diversity variables. Tree species diversity variables had clear positive correlations with stem density (Species richness: \ccsi{}; Shannon equitability: \ccei{}).

@@

Cluster 2, which contained vegetation dominated by \textit{Julbernadia paniculata} showed no significant effect of tree species diversity on AGB (\strucbsb{}), in contrast to the other vegetation clusters, and had a smaller positive effect of structural diversity on AGB (\strucbhb{}) (\autoref{struc_model_slopes_all}).

@@

We repeatedly sub-sampled the plot dataset to build \subn{} datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density. Each dataset consisted of approximately \subp{} plots with overlap of plot identity between subsampled datasets. The same SEM specification was used as above (\autoref{struc_mod}) except that paths via stem density were removed.  \autoref{sem_struc_stems_ha} shows a positive effect of tree species diversity on AGB as stem density increases. There appears to be a minimum stem density threshold at \textapprox{}295 stems ha\textsuperscript{-1} below which there appears to be no effect of tree diversity on biomass. The effect of structural diversity on AGB appears to decrease with increasing stem density, though the relationship is less clear.

@@

Similar to the model which only considered tree species and structural diversity, the direct effect of species diversity on structural diversity was positive, while structural diversity itself had a positive effect on AGB (\rghb{}).

@@

Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass (\autoref{int_plots}). The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant (\moderp{}), while the interaction term of tree species diversity and soil fertility was marginally significant (\moders{}). As moisture availability increased, the effect of tree species diversity on AGB became stronger. See \autoref{mois_div_int_mod} and \autoref{soil_div_int_mod} for full regression tables.

@@

\bibliography{/Users/johngodlee/google_drive/bib/papers/lib.bib}
